\chapter{CONCLUSION}
\label{chap:conclusion}

The MediVault project successfully demonstrates that blockchain technology can be a transformative force in healthcare data management. By fundamentally re-architecting the data ownership model to be patient-centric, MediVault restores the much-needed elements of trust and transparency to the process of medical data sharing. The system effectively proves that a decentralized approach can overcome many of the security and interoperability limitations inherent in traditional, centralized EHR systems.

The addition of the Anonymized Opt-in Research Module strengthens this system further by introducing an ethical and privacy-preserving framework for medical research. In MediVault, patients are granted dual layers of control: one over the sharing of their personal health records with specific providers, and another over their participation in broader research initiatives. This dual-consent architecture represents a fundamental shift in healthcare technology, transforming patients from passive data sources into empowered stakeholders who are active participants in their own healthcare journey.

Several avenues exist for future improvements. A significant enhancement would be the integration of IPFS (InterPlanetary File System) for decentralized file storage, which would remove reliance on any form of centralized storage for encrypted data blobs. Deploying the smart contracts on public Ethereum testnets, such as Sepolia or Goerli, would allow for more realistic, large-scale testing. The frontend could also be expanded into a more comprehensive multi-role interface that provides distinct dashboards and functionalities for doctors, patients, and researchers.

Ultimately, MediVault achieves what centralized healthcare systems have struggled to provide: security, transparency, patient control, and ethical data sharing within a single, unified platform. Through this innovation, patients are no longer mere subjects of data collection; they become active contributors to medical advancement, all while remaining fully in control of their personal privacy.
