\documentclass[12pt, a4paper]{report}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{geometry}
\geometry{a4paper, margin=1in}
\usepackage{setspace}
\onehalfspacing
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=cyan,
}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}
\usepackage[style=ieee, backend=biber]{biblatex}
\addbibresource{references.bib}

\begin{document}

% --- TITLE PAGE ---
\begin{titlepage}
    \centering
    \vspace*{1cm}
    {\Huge \bfseries \textit{🌿 MediVault — A Blockchain-Based Healthcare Data Sharing System}}

    \vspace{1.5cm}

    {\Large \bfseries A Project Report}

    \vspace{1cm}

    {\large Submitted in partial fulfillment of the requirements for the degree of}

    \vspace{0.5cm}

    {\Large \bfseries Bachelor of Technology}

    \vspace{0.5cm}

    {\large in}

    \vspace{0.5cm}

    {\Large \bfseries Computer Science and Engineering}

    \vfill

    {\large \bfseries Submitted by:} \\
    {\Large \bfseries Pratiksha} \\
    {\large (Roll No.: 220101098)}

    \vspace{1.5cm}

    {\large \bfseries Under the supervision of:} \\
    {\Large \bfseries Dr. Sanasam Chanu Inunganbi} \\
    {\large Assistant Professor, CSE} \\
    {\large Email: inunganbi[at]iiitmanipur[dot]ac[dot]in} \\
    {\large Phone: +91 70058 44201}

    \vspace{2cm}

    {\large \bfseries Department of Computer Science and Engineering} \\
    {\large \bfseries Indian Institute of Information Technology, Manipur} \\
    {\large May 2024}

\end{titlepage}

% --- DECLARATION ---
\chapter*{DECLARATION}
\addcontentsline{toc}{chapter}{DECLARATION}
I hereby declare that the project report entitled “\textbf{MediVault — A Blockchain-Based Healthcare Data Sharing System}” submitted by me to the Department of Computer Science and Engineering, Indian Institute of Information Technology, Manipur, is a bonafide and authentic work carried out by me under the supervision of Dr. Sanasam Chanu Inunganbi. The work presented here is original and has not been submitted in part or full to any other University or Institute for the award of any degree or diploma.

\vspace{2cm}
\noindent
\textbf{Pratiksha} \\
(220101098)
\newpage

% --- CERTIFICATE ---
\chapter*{CERTIFICATE}
\addcontentsline{toc}{chapter}{CERTIFICATE}
This is to certify that the project report entitled “\textbf{MediVault — A Blockchain-Based Healthcare Data Sharing System}” is a bonafide work carried out by \textbf{Pratiksha (220101098)} in partial fulfillment of the requirements for the award of the degree of Bachelor of Technology in Computer Science and Engineering at the Indian Institute of Information Technology, Manipur. The work has been carried out under my guidance and is to the best of my knowledge, an original piece of work.

\vspace{2cm}
\noindent
\textbf{Dr. Sanasam Chanu Inunganbi} \\
Assistant Professor, CSE \\
Supervisor
\newpage

% --- ACKNOWLEDGEMENT ---
\chapter*{ACKNOWLEDGEMENT}
\addcontentsline{toc}{chapter}{ACKNOWLEDGEMENT}
I would like to express my deepest gratitude to my project supervisor, \textbf{Dr. Sanasam Chanu Inunganbi}, for her invaluable guidance, constant encouragement, and insightful feedback throughout the duration of this project. Her expertise and vision were instrumental in shaping the direction of this work.

I am also thankful to the faculty and staff of the Department of Computer Science and Engineering for providing the necessary infrastructure and a conducive environment for research and learning.

Finally, I would like to thank my family and friends for their unwavering support and encouragement, which helped me navigate the challenges of this project.

\vspace{2cm}
\noindent
\textbf{Pratiksha}
\newpage

% --- ABSTRACT ---
\chapter*{ABSTRACT}
\addcontentsline{toc}{chapter}{ABSTRACT}
Traditional Electronic Health Record (EHR) systems are plagued by issues of data security, interoperability, and patient privacy. Centralized architectures are vulnerable to breaches, and patients often lack control over their own medical information. This project, \textbf{MediVault}, introduces a blockchain-based healthcare data sharing system designed to address these critical challenges. By leveraging Ethereum, Solidity smart contracts, and a decentralized architecture, MediVault empowers patients with full ownership and control over their encrypted medical records.

The system facilitates secure and transparent data exchange between patients, doctors, and researchers. A key innovation of this project is the \textbf{Anonymized Opt-in Research Module}, which allows patients to voluntarily contribute de-identified data for scientific research while maintaining full, revocable control over their consent. This feature creates an ethical framework for large-scale data analysis without compromising individual privacy. The implementation demonstrates a robust, patient-centric model that ensures data integrity, security, and accountability, paving the way for a more transparent and ethical healthcare ecosystem.
\newpage

% --- TABLE OF CONTENTS ---
\tableofcontents
\newpage

% --- CHAPTERS ---
\input{introduction.tex}
\input{literature_survey.tex}
\input{requirement_engineering.tex}
\input{system_design.tex}
\input{implementation.tex}
\input{results.tex}
\input{conclusion.tex}

\printbibliography

\end{document}
